% !TEX root = ../CedricDe Schepper2023_Thesis.tex

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

I would like to express my gratitude to several people for supporting me throughout the writing of this master's thesis. First, I would like to thank Prof. Dr. Serge Demeyer for giving me the opportunity to perform research in this interesting and relevant subject. A special thanks goes to Joey De Pauw, whose continuous feedback and support was crucial to finishing my thesis. His support ranged from giving invaluable suggestions, to helping me stay on track, to proofreading every intermediate version. 

Additionally, I want to thank Annelies Barentsen, Heidi Snellings, and Annick Van Son from the administration of the \acrlong{ua}. During several meetings, they shared their expert knowledge into the difficulties and requirements they faced when creating the necessary exam schedules. Their insights were essential to implementing my timetabling algorithm and interpreting the results.

Furthermore, I couldn't have done this without the support of Bram De Schouwer and Niels Boone. Completing this thesis combined with working full-time at Deloitte has not been easy on me. Because of this, I'm very grateful for their support, especially for giving me the flexibility to take some days off when needed so I could focus fully on my research. 

Finally, I'd like to thank my family and friends for their never-ending support. Especially my parents and my sister Aline for supporting me during my entire academical journey. Also Jolien, for helping me stay motivated and always being there to encourage me.
